---
title: "NFL Draft"
author: "InfiniteFlash"
date: "2023-02-11"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
    df_print: paged
---

The purpose of this document to provide a template of how to evaluate draft trades given prior trade precedent set forth before. The data I use in this document originally comes from this link: https://github.com/nflverse/nfldata/blob/master/DATASETS.md

# Data Import

### Import packages

Here are some basic packages to help wrangle & manipulate data.

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(sqldf)
library(kableExtra)
library(ggthemes)
```

### Import tables

Below are some important draft related tables I use to understand the draft's history & trades.

```{r, warning = FALSE, message= FALSE}
# Import draft tables 
trades <- read_csv("https://raw.githubusercontent.com/leesharpe/nfldata/master/data/trades.csv")
draft_picks <- read_csv("https://raw.githubusercontent.com/leesharpe/nfldata/master/data/draft_picks.csv")
draft_values <- read_csv("https://raw.githubusercontent.com/leesharpe/nfldata/master/data/draft_values.csv")

# Idenfitifes whether a player is a QB or not
draft_picks%>%
select(pfr_id, position)%>%
filter(!is.na(pfr_id), position == 'QB')%>%
unique() -> pfr_positions
```

\newpage

### Trades 

```{r}
trades%>%
head()%>%
kable()%>%
kable_classic(full_width = F, html_font = "Cambria")%>%
kable_styling(latex_options = 'hold_position')
```

The `trades` table contains all players and draft picks involved in all trades since 2002.

### Draft selections 

```{r}
draft_picks%>%
head()%>%
kable()%>%
kable_classic(full_width = F, html_font = "Cambria")%>%
kable_styling(latex_options = 'hold_position')
```

The `draft_picks` table contains all draft pick selections since 1980.

### Pick values 

```{r}
draft_values%>%
head()%>%
kable()%>%
kable_classic(full_width = F, html_font = "Cambria")%>%
kable_styling(latex_options = 'hold_position')
```

The `draft_values` table provides the estimated value contained within a given draft pick. 


\newpage

# Table creation

So we've got 3 tables of historical draft history & values: now what?

Let's define some criteria:

- Trades involving players are difficult to quantity. For the sake of this project, I'm going to ignore all trades that involve 1 or more players. 
- Future draft picks are valued a full round later than the current draft

```{r, warning = FALSE, message= FALSE, cache=TRUE}

# Building out a base table

trade_value_base <- trades%>%
group_by(trade_id)%>% 
mutate(n = n(), draft_picks = ifelse(any(is.na(pick_season)), 0, n()))%>% 
ungroup()%>%
#future picks are evaluated a round later
mutate(draft_pick_EV = ifelse(season != pick_season, (pick_round-1)*32+32, pick_number))%>% 
#filter out trades with players
filter(n == draft_picks)%>% 
left_join(., draft_values, by = c("draft_pick_EV"="pick"))%>%
left_join(., pfr_positions, by = c("pfr_id"="pfr_id"))%>%
select(trade_id: pick_number, position, draft_pick_EV:pff)%>%
mutate(position = ifelse(position == 'QB', 'QB', 'non-QB'))


# Trade value given & received in all trades & value chart types

trade_value_all <- trade_value_base%>%
group_by(trade_id, gave, received)%>%
summarise(stuart = sum(stuart),
          johnson = sum(johnson),
          hill = sum(hill),
          otc = sum(otc),
          pff = sum(pff))%>%
gather(chart_type, amount, stuart:pff)

# Gets information about the top pick traded in a given trade
trade_value_base%>%
group_by(trade_id)%>%
filter(draft_pick_EV == min(draft_pick_EV))%>%
select(trade_id, season, trade_date, gave, received, pick_number) -> information

# Picks traded by each team concatenated
trade_value_base%>%
    group_by(trade_id, gave, received)%>%
    summarise(trade_results = paste(sort(pick_number), collapse = ', ')) -> picks


#Write tables out to sql. Not necessary as the SQL query below can be run via R code.
con <- DBI::dbConnect(RSQLite::SQLite(), dbname = "NFL_draft.sqlite")
dbWriteTable(con,"trade_value_all", trade_value_all, overwrite = TRUE)
dbWriteTable(con,"trade_value_base", trade_value_base, overwrite = TRUE)
dbWriteTable(con,"information", information, overwrite = TRUE)
dbWriteTable(con,"picks", picks, overwrite = TRUE)
dbWriteTable(con,"draft_picks", draft_picks, overwrite = TRUE)

```

\newpage

### Build out summary set

Below is combining all of the information above together.

```{r, warning = FALSE, message= FALSE, cache=TRUE}
query = "select a.trade_id, a.gave, a.received, a.chart_type, 
         a.amount as amount_given,
         b.amount as amount_received,
         a.amount-b.amount as difference,
         (a.amount-b.amount)/a.amount as given_percentage,
         (a.amount-b.amount)/b.amount as received_percentage,
         c.season, c.trade_date, c.pick_number, 
         case when f.position = 'QB' then 'QB' 
         else 'non-QB' end as position,
         d.trade_results as given,
         e.trade_results as offered,
         f.pfr_name
         from trade_value_all as a
         left join trade_value_all as b on a.trade_id = b.trade_id and a.chart_type = b.chart_type and a.gave != b.gave
         inner join information as c on a.trade_id = c.trade_id and a.gave = c.gave and a.received = c.received
         inner join picks as d on a.trade_id = d.trade_id and a.gave = d.gave
         inner join picks as e on a.trade_id = e.trade_id and a.gave = e.received
         left join draft_picks as f on c.season = f.season and c.pick_number = f.pick
         where b.amount > 0 and a.amount > 0
         order by 1, a.chart_type
        "
summary_set <- sqldf(query)

summary_set%>%
head()%>%
kable()%>%
kable_classic(full_width = F, html_font = "Cambria")%>%
kable_styling(latex_options = 'hold_position') 

```




A few notes:

- `amount_given` is the amount that the **gave** team traded away.
- `amount_received` is the amount that the **gave** team received.
- `difference` is the expected amount of value the **gave** team received.
- `given_percentage` and `received_percentage` give an idea of how much value the **gave** and **received** teams got relative to the amount trade away.





```{r,echo=FALSE}
# write.csv(summary_set, 'draft_trade_evals.csv')
# summary_set <- read.csv('draft_trade_evals.csv')
```


# Results: Draft history value

```{r, warning = FALSE, message= FALSE, cache=TRUE}
summary_set%>%
mutate(chart_type = tools::toTitleCase(chart_type))%>%
filter(pick_number <= 32, amount_received > 0, amount_given > 0
       # , position %in% c('QB', 'non-QB')
       # ,chart_type == 'hill'
       )%>%
# select(-position)%>%
ggplot(aes(x = pick_number, y= amount_received/amount_given))+
geom_point(size = 1.5, aes(size = 2, color = position, shape = position))+
geom_vline(xintercept = 1)+
geom_hline(yintercept = 1)+
geom_smooth(se= FALSE,
            color = 'black')+
theme_wsj()+
facet_wrap(. ~ chart_type, scale = 'free')+
labs(x = '\nPick #',  y= 'Expected % received\n', title = 'Pick # vs Draft value received')+
        # theme(legend.position = c(0.85, 0.25), 
        #       legend.background = element_rect(fill = "white", colour = NA),
        #       text = element_text(size=10),
        #       plot.title = element_text(size = 15),
        #       axis.title=element_text(size=15),
        #       legend.title = element_text(size = 10))+
  
  theme(legend.position = c(0.85, 0.25),
legend.background = element_rect(fill = "white", colour = NA),
  plot.title = element_text(size = 15),
text = element_text(size=10),
axis.title=element_text(size=20),
strip.text = element_text(size = 11),
legend.title = element_blank())+
 scale_y_continuous(labels = scales::percent)
```

And guala, now we have a historical draft trade chart showing how much teams typically recoup when trading down. 

`Non-QB` and `QB` represents if team trading up, picked a QB with the top pick being traded.

It is nice to see the old trope of teams overpaying, when they pay for QB. Hence the cluster of QBs in the top left corner of each subplot.


```{r, cache=TRUE, echo=FALSE, eval=FALSE}
summary_set%>%
filter(pick_number <= 16, amount_received > 0, amount_given > 0
       , position != 'non-QB'
       )%>%
ggplot(aes(x = pick_number, y= amount_received/amount_given, color = position))+
geom_point(size = 1.5)+
geom_vline(xintercept = 1)+
geom_hline(yintercept = 1)+
# geom_smooth(se= FALSE, color = 'black')+
theme_bw()+
facet_wrap(. ~ chart_type, scale = 'free')+
labs(x = '\nPick #', 
     y = 'Additional amount received in return\n', 
     title = 'Pick # (since 2011) vs the expected amount received in return.',
     caption = '')+ 
        theme(legend.position = c(0.85, 0.25), 
              legend.background = element_rect(fill = "white", colour = NA),
              text = element_text(size=20),
              axis.title=element_text(size=30),
              legend.key.size = unit(0.1, "cm")),
              strip.text = element_text(size = 20))
    theme(legend. title = element_blank()
```


<!-- # Minimum references -->

<!-- ```{r, cache=TRUE} -->
<!-- summary_set%>% -->
<!-- filter(pick_number <= 5, amount_received > 0, amount_given > 0, position == 'QB')%>% -->
<!-- group_by(chart_type)%>% -->
<!-- summarise(minimum = min(amount_received-amount_given)) -->
<!-- ``` -->

<!-- Based on this chart, I think it's reasonable to draw that teams trading down recoup most of the value. -->

<!-- # Trade with Indy -->

<!-- ```{r} -->
<!-- draft_values%>% -->
<!-- filter(pick %in% c(1, #Bears picks -->
<!--                    4, 35, 79, 105, 138, 162, 222, 237, # Colts picks this year -->
<!--                    16, 48, 80, 112, 144, 176 #Expected picks next year -->
<!--                    )) -->
<!-- ``` -->

<!-- Based on the chart above, the expected minimum from a trade with the Colts would be: -->

<!-- Trade Chart Type: **Expected Minimum Points** -->

<!-- - Stuart: **`r 34.6 + 16.5` points** -->

<!--      - 2023 1st (#4), 2023 3rd (#79), 2023 6th (#176) -->
<!--      - 2024 1st (#16)  -->

<!-- equals **`r 25.8+16.9+6.8+1.7`** -->

<!-- - Johnson: **`r 3000+117` points** -->


<!--      - 2023 1st (#4), 2023 3rd (#79), 2023 4th (#105), 2023 5th (#138) -->
<!--      - 2024 1st (#16)  -->

<!-- equals **`r 1800+1000+195+84+37`** -->

<!-- - Hill: **`r 1000-44` points** -->

<!--      - 2023 1st (#4), 2023 2nd (#79) -->
<!--      - 2024 1st (#16)  -->

<!-- equals **`r 490.52+305.41+170.37`** -->

<!-- - OTC: **`r 3000+1784` points** -->

<!--      - 2023 1st (#4), 2023 2nd (#79) -->
<!--      - 2024 1st (#16)  -->

<!-- equals **`r 2270+1540+1128`** -->

<!-- - PFF: **`r 1.135+.674` points** -->

<!--      - 2023 1st (#4), 2023 5th (#144) -->
<!--      - 2024 1st (#16)  -->

<!-- equals **`r 1.027+.66+.134`** -->


<!-- ```{r, message=FALSE, warning=FALSE} -->
<!-- library(janitor) -->
<!-- library(tools) -->

<!-- draft_values[c(4, 35, 48, 48, 105), ]%>% -->
<!-- mutate(pick = pick%>%as.character())%>% -->
<!-- adorn_totals("row")%>% -->
<!-- bind_rows(draft_values[c(1), ]%>%mutate(pick = pick%>%as.character()))%>% -->
<!-- bind_rows(data.frame(pick = 'Minimum', stuart = 34.6+16.5, johnson = 3117, hill= 856, otc = 4784, pff = 1.809))%>% -->
<!-- arrange(pick)%>% -->
<!-- mutate(order = c(1,6 ,3 , 2, 4, 5, 7, 8))%>% -->
<!-- arrange(order)%>% -->
<!-- mutate(pick = ifelse(pick == 'Total', 'Discord suggested offer', pick))%>% -->
<!-- magrittr::set_colnames(toTitleCase(colnames(.)))%>% -->
<!-- kable() -->
<!-- ``` -->

